---
title: "DATA 301 Assignment IV"
author: "Oliver Donaldson, 300659923"
date: "22 September 2026"
format:
  pdf:
    pdf-engine: xelatex
    geometry: margin=2.4cm
    fig-align: center
    fig-width: 7
    fig-height: 3.4
    fig-pos: "H"
    df-print: default
    tbl-cap-location: top
    include-in-header:
      text: |
        \usepackage{booktabs}
        % needspace.sty's \needspace, inlined: page-break before a table rather than split it
        \makeatletter
        \newcommand{\needspace}[1]{\begingroup\setlength{\dimen@}{#1}\vskip\z@\@plus\dimen@\penalty -100\vskip\z@\@plus -\dimen@\vskip\dimen@\penalty 9999\vskip -\dimen@\endgroup}
        \makeatother
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", out.width = "95%")
suppressPackageStartupMessages({library(fpp2); library(ggplot2); library(knitr)})
theme_set(theme_bw())   # white background: prints and photocopies legibly
R_REPS <- 50000   # Monte Carlo replications, justified in 1.3; the render takes ~4 min
B_BOOT <- 200     # bootstrap replications, fixed by the assignment
```

```{r}
#| label: exercise-1
# index j and interpolation weight g of a sample quartile, quantile type 8 or 7
qidx <- function(n, p, type = 8) {
  h <- if (type == 8) (n + 1/3) * p + 1/3 else (n - 1) * p + 1
  j <- floor(h)
  list(j = j, g = h - j)
}

lam_ml <- function(xs) 1 / mean(xs)

lam_iqr <- function(xs, type = 8) {          # xs must be sorted
  n <- length(xs)
  a <- qidx(n, .25, type)
  b <- qidx(n, .75, type)
  lo <- xs[a$j] + a$g * (xs[a$j + 1] - xs[a$j])
  hi <- xs[b$j] + b$g * (xs[b$j + 1] - xs[b$j])
  log(3) / (hi - lo)
}

# B bootstrap replicates of lam_iqr at once: resample indices of the sorted sample and
# find each resample's order statistics by a counting sort
boot_iqr <- function(xs, B) {
  n   <- length(xs)
  idx <- sample.int(n, n * B, replace = TRUE)
  off <- rep((0:(B - 1)) * n, each = n)      # row b occupies slots (b-1)n+1 .. bn
  cnt <- matrix(tabulate(off + idx, nbins = n * B), nrow = B, byrow = TRUE)
  cum <- t(apply(cnt, 1, cumsum))
  pk  <- function(k) xs[max.col(cum >= k, ties.method = "first")]
  a <- qidx(n, .25)
  b <- qidx(n, .75)
  lo <- pk(a$j) + a$g * (pk(a$j + 1) - pk(a$j))
  hi <- pk(b$j) + b$g * (pk(b$j + 1) - pk(b$j))
  log(3) / (hi - lo)
}

# first-order bootstrap bias correction; resamples with zero IQR are left out
lam_bc <- function(xs, B = B_BOOT) {
  hat  <- lam_iqr(xs)
  star <- boot_iqr(xs, B)
  ok   <- is.finite(star)
  c(est = 2 * hat - mean(star[ok]), bad = sum(!ok))
}

run_cell <- function(lambda, n, R, B = B_BOOT) {
  ml <- i8 <- i7 <- bc <- numeric(R); nbad <- 0L
  for (r in seq_len(R)) {
    xs <- sort(rexp(n, lambda))
    ml[r] <- lam_ml(xs); i8[r] <- lam_iqr(xs, 8); i7[r] <- lam_iqr(xs, 7)
    o <- lam_bc(xs, B); bc[r] <- o[["est"]]; nbad <- nbad + o[["bad"]]
  }
  st <- function(e) { se <- (e - lambda)^2
    c(mse = mean(se), mcse = sd(se) / sqrt(R), top1 = max(se) / sum(se)) }
  list(lambda = lambda, n = n, p_neg = mean(bc < 0),
       ML = st(ml), IQR8 = st(i8), IQR7 = st(i7), BC = st(bc),
       bad_rate = nbad / (R * B))
}

set.seed(20260922)
grid <- expand.grid(n = c(5, 30, 100), lambda = c(.1, 1, 10))
res  <- Map(function(l, n) run_cell(l, n, R_REPS), grid$lambda, grid$n)

# values quoted in the text
cell <- function(l, n) res[[which(grid$lambda == l & grid$n == n)]]
stat <- function(est, l, n, s = "mse") cell(l, n)[[est]][[s]]
pc   <- function(x) sprintf("%.0f", 100 * x)
f1   <- function(x) sprintf("%.1f", x)
big  <- function(x) format(signif(x, 3), big.mark = ",", scientific = FALSE)
lams <- c(.1, 1, 10)

wx   <- c(0, 1, 1, 1, 2, 2, 2, 4, 5, 8)      # Wicklin's ten-point example
wiqr <- sapply(1:9, function(k) diff(quantile(wx, c(.25, .75), type = k, names = FALSE)))

rel   <- unlist(lapply(res[grid$n > 5], function(z)
  sapply(z[c("ML", "IQR8", "BC")], function(s) s[["mcse"]] / s[["mse"]])))
exact <- function(l, n) l^2 * (n + 2) / ((n - 1) * (n - 2))
z_ml  <- max(abs(sapply(res, function(z) (z$ML[["mse"]] - exact(z$lambda, z$n)) / z$ML[["mcse"]])))
# sd / mean of the ML squared error at n = 5, from E[lam_ml^k] = n^k Gamma(n - k) / Gamma(n), lambda = 1
mom   <- function(k, n = 5) n^k * gamma(n - k) / gamma(n)
rsd5  <- sqrt(mom(4) - 4 * mom(3) + 6 * mom(2) - 4 * mom(1) + 1 - exact(1, 5)^2) / exact(1, 5)
ratio <- sapply(res[grid$n > 5], function(z) z$IQR8[["mse"]] / z$ML[["mse"]])
top_bc <- sapply(lams, function(l) stat("BC", l, 5, "top1"))
```

# Exercise 1

## 1.1 Order statistics [10%]

**Wicklin's analysis.** Wicklin (2017) compares the nine sample-quantile definitions catalogued
by Hyndman & Fan (1996). All nine take two adjacent order statistics $x_{(j)}, x_{(j+1)}$, with
$j = \lfloor np + m \rfloor$, and return $(1-\gamma)x_{(j)} + \gamma x_{(j+1)}$; they differ only in
the offset $m$ and in how the weight $\gamma$ is set. Types 1--3 round, returning a data value or
the average of two; types 4--9 interpolate linearly and can return any value in the range of the
data. SAS implements five of them and defaults to type 2; R implements all nine and defaults to
type 7. Plotting all nine for a ten-point sample, Wicklin finds that they differ most for small
samples, near large gaps between data values and in the tails, and hardly at all for 100 or more
observations away from gaps. His conclusion is that the choice matters for small samples, so the
definition used should be reported; otherwise most analysts accept their software's default. On
his ten-point sample the nine sample IQRs range from `r sprintf("%.2f", min(wiqr))` to
`r sprintf("%.2f", max(wiqr))`.

**Which type I would use: type 8.** Of Hyndman & Fan's (1996) six desirable properties, only type 5
has all six, but they call it a compromise that no estimation argument justifies. Types 6--9 have
five each, and of these they recommend type 8 because it is approximately median-unbiased whatever
the underlying distribution, which suits an estimator whose purpose is not to rely on the
exponential model. The type also matters for this
estimator, which is the reciprocal of an IQR and so is driven by samples with a small IQR. At
$n = 5$, type 7 gives $\text{IQR} = X_{(4)} - X_{(2)}$, the sum of two independent exponential
spacings. Its density rises only linearly from zero, so $\mathbb{E}[\text{IQR}^{-2}] = \infty$ and
type-7 $\widehat\lambda_{\mathrm{IQR}}$ has infinite MSE. In the simulation of Section 1.3 its
estimated $\mathrm{MSE}/\lambda^2$ comes out as
`r paste(round(sapply(lams, function(l) stat("IQR7", l, 5) / l^2)), collapse = ", ")` at the three
values of $\lambda$ instead of being equal, and a single sample supplies up to
`r pc(max(sapply(lams, function(l) stat("IQR7", l, 5, "top1"))))`% of it. Type 8 interpolates
between $X_{(1)}, X_{(2)}$ and $X_{(4)}, X_{(5)}$, so its IQR is a weighted sum of four spacings and
its MSE is finite. At $n = 30$ and $n = 100$ type 7 still has
`r pc(stat("IQR7", 1, 30) / stat("IQR8", 1, 30) - 1)`% and
`r pc(stat("IQR7", 1, 100) / stat("IQR8", 1, 100) - 1)`% more MSE than type 8.

**Default.** R `r getRversion()` uses type `r formals(stats:::quantile.default)$type` (in
`quantile()` and `IQR()`). All computations here set type 8 explicitly.

## 1.2 Bootstrap bias-corrected estimator [20%]

The bias $b(\lambda) = \mathbb{E}_\lambda[\widehat\lambda_{\mathrm{IQR}}] - \lambda$ depends on the
unknown distribution $F$. The bootstrap replaces $F$ by the empirical distribution
$\widehat F_n$ of the observed sample $\mathbf x$ (Robert & Casella, 2009, §1.5), which gives the
bias estimate

$$\widehat b = \mathbb{E}_*\big[\widehat\lambda_{\mathrm{IQR}}(\mathbf X^*)\big]
  - \widehat\lambda_{\mathrm{IQR}}(\mathbf x)
  \approx \frac1B\sum_{b=1}^{B} \widehat\lambda^*_b - \widehat\lambda_{\mathrm{IQR}}(\mathbf x),
\qquad \widehat\lambda^*_b = \widehat\lambda_{\mathrm{IQR}}(\mathbf x^*_b),$$

where $\mathbf x^*_1, \dots, \mathbf x^*_B$ are $B = 200$ resamples of $\mathbf x$ drawn with
replacement. Subtracting it gives the first-order bias-corrected estimator

$$\widetilde\lambda = \widehat\lambda_{\mathrm{IQR}} - \widehat b
  = 2\,\widehat\lambda_{\mathrm{IQR}} - \frac1B\sum_{b=1}^{B}\widehat\lambda^*_b .$$

It is first-order because it subtracts a single bootstrap estimate of the bias, without iterating
the correction.

**Implementation** (`lam_iqr`, `boot_iqr` and `lam_bc` in the `.qmd`). The type-8 quartiles are
computed directly from the order statistics, and all $B$ resamples are drawn and ordered at once. A
resample with zero IQR gives $\widehat\lambda^* = \infty$; at $n = 5$ this happens only when all five
values coincide (probability $5^{-4} = 1/625$). Such resamples are left out of the average.

## 1.3 Monte Carlo experiment [20%]

Only $\widehat\lambda_{\mathrm{ML}}$ has a simple closed-form MSE; $\widetilde\lambda$ depends on
the bootstrap resamples and has none, so the comparison is made by simulation. In each of the nine cells $(\lambda, n) \in \{0.1, 1, 10\} \times \{5, 30, 100\}$, $R$ samples are
drawn from $\text{Exp}(\lambda)$ and the three estimators computed on each. The MSE is estimated by
the average squared error over the $R$ samples, with Monte Carlo standard error $s/\sqrt R$
(Robert & Casella, 2009, §3.2).

**Number of replications.**

- *For each $\lambda$.* All three estimators satisfy
  $\widehat\lambda(c\mathbf x) = \widehat\lambda(\mathbf x)/c$, and $X \sim \text{Exp}(\lambda)$ is
  $Z/\lambda$ with $Z \sim \text{Exp}(1)$, so $\mathrm{MSE}(\lambda, n) = \lambda^2\,\mathrm{MSE}(1, n)$.
  The *relative* Monte Carlo error therefore does not depend on $\lambda$, and the same $R$ serves
  every $\lambda$ at a given $n$.
- *For each $n$.* Aiming for a relative standard error of 1%, a pilot with $R = 2{,}000$ at
  $\lambda = 1$ (`working/01-pilot-mc-error.R`) gave at most 4.7% for $n = 30$ and 3.8% for
  $n = 100$, so $n = 30$ needs $R \approx 2000 \times 4.7^2 \approx 44{,}000$. I use $R = 50{,}000$
  in every cell; the achieved relative errors at $n \ge 30$ are
  `r f1(100 * min(rel))`--`r f1(100 * max(rel))`% (Table 1).
- *At $n = 5$.* $s/\sqrt R$ is only valid when the squared error has finite variance (Robert &
  Casella, 2009, §3.2). For $\widehat\lambda_{\mathrm{ML}}$ it does (the fourth moment is finite for
  $n > 4$), but its exact moments give the squared error a standard deviation `r f1(rsd5)` times
  its mean, so 1% would need about `r big(1e4 * rsd5^2)` replications; its MSE is also known
  exactly, so the simulation only has to check the code. For $\widehat\lambda_{\mathrm{IQR}}$ the squared error has
  infinite variance at $n = 5$, and $\widetilde\lambda$ has no finite MSE at all (below), so no $R$
  can deliver a 1% error. $R = 50{,}000$ is therefore kept, and the $n = 5$ standard errors are
  reported but not relied on.

\needspace{17\baselineskip}

```{r}
#| label: tbl-main
#| tbl-cap: "Monte Carlo MSE, with relative Monte Carlo standard error in parentheses; $R = 50{,}000$, $B = 200$. At $n = 5$ the $\\widetilde\\lambda$ entries do not estimate a finite quantity."
fmt <- function(z) sprintf("%s (%.1f%%)", big(z[["mse"]]), 100 * z[["mcse"]] / z[["mse"]])
kable(data.frame(
  lambda = sapply(res, `[[`, "lambda"), n = sapply(res, `[[`, "n"),
  ML = sapply(res, function(z) fmt(z$ML)), IQR = sapply(res, function(z) fmt(z$IQR8)),
  tilde = sapply(res, function(z) fmt(z$BC))),
  col.names = c("$\\lambda$", "$n$", "$\\widehat\\lambda_{\\mathrm{ML}}$",
                "$\\widehat\\lambda_{\\mathrm{IQR}}$", "$\\widetilde\\lambda$"),
  booktabs = TRUE, escape = FALSE, align = "rrrrr", row.names = FALSE)
```

**Results.**

- $\widehat\lambda_{\mathrm{ML}}$ has the smallest MSE in every cell, `r f1(min(ratio))`--`r f1(max(ratio))`
  times smaller than $\widehat\lambda_{\mathrm{IQR}}$ for $n \ge 30$, as expected when the model is
  correct. Its simulated MSE agrees with the exact value $\lambda^2(n+2)/[(n-1)(n-2)]$ to within
  `r f1(z_ml)` standard errors in every cell, which checks the simulation code.
- The bias correction increases the MSE. $\widetilde\lambda$ has
  `r pc(stat("BC", 1, 30) / stat("IQR8", 1, 30) - 1)`% more MSE than $\widehat\lambda_{\mathrm{IQR}}$
  at $n = 30$ and `r pc(stat("BC", 1, 100) / stat("IQR8", 1, 100) - 1)`% more at $n = 100$ (at
  every $\lambda$).
- At $n = 5$, $\widetilde\lambda$ has no finite MSE. If $a < b$ are adjacent sample values, the
  resample $\{a, b, b, b, b\}$ (probability 1/625) has type-8 IQR $(b - a)/3$ and contributes
  $\widehat\lambda^* = 3\ln 3/(b - a)$. As $b - a$ is an exponential spacing,
  $\mathbb{E}[1/(b - a)] = \infty$, so $\widetilde\lambda$ has expectation $-\infty$. The Table 1
  entries for $\widetilde\lambda$ at $n = 5$ are therefore not estimates of anything:
  $\mathrm{MSE}/\lambda^2$ comes out as
  `r paste(big(sapply(lams, function(l) stat("BC", l, 5) / l^2)), collapse = "; ")` at the three
  values of $\lambda$, one sample supplies `r pc(min(top_bc))`--`r pc(max(top_bc))`% of each
  total, and `r pc(mean(sapply(lams, function(l) cell(l, 5)$p_neg)))`% of the estimates are
  negative.

**Conclusion.** $\widehat\lambda_{\mathrm{ML}}$ is the best of the three throughout. The bootstrap
correction does not improve $\widehat\lambda_{\mathrm{IQR}}$ in MSE at any sample size, and at
$n = 5$ it breaks down completely.

\newpage

# Exercise 2

```{r}
#| label: exercise-2
full <- window(ausbeer, start = c(1956, 1), end = c(2009, 4))   # complete years only
m    <- matrix(full, ncol = 4, byrow = TRUE)                     # one row per year
yr   <- 1956:2009
ann  <- rowSums(m)                                               # annual totals
inc  <- setNames(diff(ann), yr[-1])                              # change on the previous year
v    <- function(y, q) m[yr == y, q]                             # one quarter's value
cm   <- function(x) format(x, big.mark = ",")                   # exact, with thousands separator
ac   <- acf(ausbeer, lag.max = 16, plot = FALSE)$acf[-1]
qm   <- round(tapply(ausbeer, cycle(ausbeer), mean))
ord  <- sum(m[, 4] > m[, 1] & m[, 1] > m[, 3] & m[, 3] > m[, 2])
swing <- tapply(m[, 4] - m[, 2], yr %/% 10, mean)                 # Q4 - Q2, by decade
level <- tapply(rowMeans(m), yr %/% 10, mean)
q4drop <- colMeans(m[yr %in% 2000:2009, ]) / colMeans(m[yr %in% 1985:1994, ]) - 1
```

```{r}
#| label: fig-time
#| fig-cap: "Quarterly beer production in megalitres, Australia, 1956 Q1 to 2010 Q2. Labelled points mark four unusual quarters."
odd <- data.frame(t  = c(1980.75, 1981, 1988.75, 1990.75), y  = c(526, 548, 598, 599),
                  lx = c(1976.5, 1985.5, 1985, 1994.5),     ly = c(370, 370, 625, 625),
                  lab = c("1980 Q4", "1981 Q1", "1988 Q4", "1990 Q4"))
autoplot(ausbeer) +
  geom_segment(data = odd, aes(lx, ly, xend = t, yend = y), colour = "grey45", linewidth = .3) +
  geom_point(data = odd, aes(t, y), colour = "red", size = 1.6) +
  geom_text(data = odd, aes(lx, ly, label = lab), size = 2.8, vjust = c(1.2, 1.2, -.3, -.3)) +
  labs(x = "Year", y = "Megalitres")
```

**1. Frequency.** The series is quarterly: `frequency(ausbeer)` is `r frequency(ausbeer)`, and it
runs from 1956 Q1 to 2010 Q2 (`r length(ausbeer)` observations). In @fig-time the pattern repeats
every four observations, with one peak and one trough a year, which is the annual cycle of
quarterly data. The data show the same period in their autocorrelation function (@fig-acf), the
plot Hyndman & Athanasopoulos (2018, §2.8) use for this series: the correlation peaks at lags 4, 8
and 12 (`r paste(sprintf("%.2f", ac[c(4, 8, 12)]), collapse = ", ")`), so each quarter is most like
the same quarter a year earlier, and dips at lags 2, 6 and 10, because the Q2 trough lies two
quarters from the Q4 peak.

```{r}
#| label: fig-acf
#| fig-cap: "Autocorrelation of quarterly beer production at lags of 1 to 16 quarters, Australia, 1956 Q1 to 2010 Q2. Dashed lines mark the approximate 95% bounds for white noise."
#| fig-height: 2.1
ggAcf(ausbeer, lag.max = 16) + labs(title = NULL)
```

**2. Overall behaviour and trend.** There is a clear long-term trend, but it is not monotone.
Annual production rises from `r cm(ann[1])` ML in 1956 to a peak of `r cm(max(ann))` ML in
`r yr[which.max(ann)]`, stays high from 1974 to 1981, falls to about 1,850 ML by 1983, recovers
briefly in 1988--90 and declines again to the mid-1990s. From 1994 to 2009 it is roughly flat
(`r cm(min(ann[yr >= 1994]))`--`r cm(max(ann[yr >= 1994]))` ML a year) with a slight downward
drift. The fall to 1983 and the 1988--90 recovery are rises and falls lasting several years with no
fixed period, so they look like cyclic rather than seasonal behaviour (Hyndman & Athanasopoulos,
2018, §2.3). The trend also shows in @fig-acf: every autocorrelation is positive and they decay only
slowly, which is how a trend appears in an ACF (Hyndman & Athanasopoulos, 2018, §2.8).

**3. Seasonality.** Yes, strong seasonality. Q4 (October--December, summer and Christmas) is the
peak and Q2 (April--June, autumn into winter) the trough. The quarter means are Q4 `r qm[4]`,
Q1 `r qm[1]`, Q3 `r qm[3]` and Q2 `r qm[2]` ML, so Q4 is about `r pc(qm[4] / qm[2] - 1)`% above Q2,
and the order Q4 > Q1 > Q3 > Q2 holds in `r ord` of the `r nrow(m)` complete years.

**4. Is the variability constant?** No. In @fig-time the seasonal swing widens as the level rises:
Q4 $-$ Q2 averages about `r round(swing[["195"]])` ML in the late 1950s and
`r round(swing[["198"]])` ML in the 1980s and 1990s. It then narrows to `r round(swing[["200"]])` ML
in the 2000s while the level falls only `r pc(1 - level[["200"]] / level[["199"]])`%, so the swing is
only partly proportional to the level. A transformation such as a log or Box--Cox transformation
could stabilise the level-related part (Hyndman & Athanasopoulos, 2018, §3.2); Question 8 argues
it is not worth doing.

**5. Unusual observations.** Each is compared with the same quarter in neighbouring years:

- 1981 Q1 (548 ML) is the largest Q1 in the series, `r v(1981, 1) - sort(m[, 1], TRUE)[2]` ML
  above the next (515 ML in 1978) and well above the Q1s either side (`r v(1980, 1)` and
  `r v(1982, 1)` ML).
- 1980 Q4 (526 ML), immediately before it, is the lowest Q4 between 1973 and 1992, against
  `r v(1979, 4)` and `r v(1981, 4)` ML either side. A weak quarter followed by a record one
  suggests production shifted from one quarter into the next.
- 1988 Q4 (598 ML) and 1990 Q4 (599 ML) are the two highest values in the series,
  `r v(1988, 4) - v(1987, 4)` and `r v(1990, 4) - v(1989, 4)` ML above the previous Q4, although
  annual production in those years (`r cm(ann[yr == 1988])` and `r cm(ann[yr == 1990])` ML) was
  below its 1981 peak.
- The step up in 1973--74, when annual production rose by `r inc[["1973"]]` ML and then
  `r inc[["1974"]]` ML, compared with
  `r min(inc[c("1970", "1971", "1972")])`--`r max(inc[c("1970", "1971", "1972")])` ML a year in
  1970--72.
- The last two observations (2010 Q1--Q2) are an incomplete year: the apparent drop at the end of
  @fig-time is the series stopping at the Q2 trough, not a decline.

```{r}
#| label: fig-season
#| fig-cap: "Quarterly beer production in megalitres by quarter and decade, Australia, complete years 1956 to 2009. Each thin line is one year; the thick line is the decade's average for each quarter."
#| fig-height: 2.9
# small multiples by decade instead of 54 overlaid years; grey and black read in print
sp  <- data.frame(q = factor(rep(paste0("Q", 1:4), nrow(m))), y = as.vector(t(m)),
                  yr = rep(yr, each = 4))
sp$dec <- paste0(sp$yr %/% 10 * 10, "s")
avg <- aggregate(y ~ q + dec, sp, mean)
ggplot(sp, aes(q, y)) +
  geom_line(aes(group = yr, colour = "One year"), linewidth = .3) +
  geom_line(data = avg, aes(group = dec, colour = "Decade average"), linewidth = .9) +
  facet_wrap(~dec, nrow = 1) +
  scale_colour_manual(NULL, values = c("One year" = "grey65", "Decade average" = "black"),
                      breaks = c("One year", "Decade average")) +
  labs(x = "Quarter", y = "Megalitres") + theme(legend.position = "bottom")
```

**6. What the seasonal plot shows.** @fig-season draws each year against the quarter, one panel per
decade. Every decade has the same shape: high in Q1, a dip to Q2, then a rise through Q3 to the Q4
peak. Only the height of the profile changes, rising from the 1950s to the 1980s and falling after.
The seasonal pattern is therefore stable while the level changes, which is hard to see in
@fig-time, where trend and season are superimposed. The panels also show the Q4 peak standing less
far above the other quarters in the 2000s than in the 1970s and 1980s.

```{r}
#| label: fig-sub
#| fig-cap: "Quarterly beer production in megalitres by quarter, Australia, 1956 Q1 to 2010 Q2. Each panel shows one quarter in time order; the horizontal line is that quarter's mean."
ggsubseriesplot(ausbeer) + labs(x = "Quarter", y = "Megalitres")
```

**7. Seasonal subseries.** Q4 has the highest typical values (mean `r qm[4]` ML) and Q2 the lowest
(`r qm[2]` ML). All four quarters follow the same rise, peak and decline, so the seasonal pattern
is broadly stable over time. The exception is Q4, which falls further after about 1990:
`r pc(-q4drop[4])`% between 1985--94 and 2000--09, against
`r pc(-max(q4drop[1:3]))`--`r pc(-min(q4drop[1:3]))`% for the other quarters, so the seasonal
amplitude shrinks over the last two decades.

```{r}
#| label: fig-log
#| fig-cap: "Quarterly beer production, Australia, 1956 Q1 to 2010 Q2, in megalitres (top) and on the log scale (bottom)."
#| fig-height: 3.8
tt <- as.numeric(time(ausbeer))
b  <- rbind(data.frame(t = tt, y = as.numeric(ausbeer),      s = "Original (ML)"),
            data.frame(t = tt, y = log(as.numeric(ausbeer)), s = "Log scale"))
b$s <- factor(b$s, levels = c("Original (ML)", "Log scale"))
ggplot(b, aes(t, y)) + geom_line(linewidth = .35) +
  facet_wrap(~s, ncol = 1, scales = "free_y") + labs(x = "Year", y = NULL)
```

**8. Detrend or transform?** *Detrend: yes.* @fig-time shows the quarterly level rising from about
260 ML in 1956 to about 500 ML around 1980 and falling back to about 430 ML, so the series has no
constant mean and is not stationary. The trend should be removed, by differencing or by modelling
it explicitly, before the rest of the structure is modelled.

*Transform: no.* A transformation is worthwhile if it makes the seasonal variation about the same
size across the whole series (Hyndman & Athanasopoulos, 2018, §3.2). @fig-log shows that the log
does not do this: on the log scale the seasonal swing is widest in the 1950s--60s and narrowest at
the end, so the log over-corrects the early growth and leaves the late narrowing in place. On the
original scale the changes in the swing are moderate and not simply proportional to the level, so I
would keep the original scale, where the values stay interpretable. As instructed, no model is
fitted.

# References

Hyndman, R. J., & Athanasopoulos, G. (2018). *Forecasting: Principles and Practice* (2nd ed.).
OTexts. <https://OTexts.com/fpp2>

Hyndman, R. J., & Fan, Y. (1996). Sample quantiles in statistical packages.
*The American Statistician*, 50(4), 361--365. <https://doi.org/10.2307/2684934>

Robert, C. P., & Casella, G. (2009). *Introducing Monte Carlo Methods with R*. Springer.

Wicklin, R. (2017). *Sample quantiles: A comparison of 9 definitions*. SAS Blogs.
<https://blogs.sas.com/content/iml/2017/05/24/definitions-sample-quantiles.html>

```{r}
#| label: session
cat(R.version.string, "|", paste("fpp2", packageVersion("fpp2")),
    "|", paste("forecast", packageVersion("forecast")),
    "|", sprintf("R = %s replications, B = %d", format(R_REPS, big.mark = ","), B_BOOT), "\n")
```
